Seagrass mapping and monitoring along the coasts of Crete, Greece

### 2011-01-23 MScDiplom_Lemenkova.tex
updated Diploma directory, added new files, renumerated Figures and Tables

\documentclass[10pt, a4paper]{article}
\usepackage{color}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[numbers]{natbib}
\usepackage{calc}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage[svgnames,table]{xcolor}
%\usepackage[table]{xcolor}
\usepackage{paralist}
\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{fancyheadings}
\topmargin = 1pt
\headheight=10mm
\headsep=7pt
%\ensuremath
\usepackage{lastpage}
\usepackage{float}
\usepackage{subfig}
\usepackage{caption}
\usepackage{appendix}
\captionsetup{margin=10pt,font=small,labelfont=bf}
\usepackage{wrapfig}
\definecolor{darkblue}{rgb}{0.0,0.0,0.3}
\setcounter{tocdepth}{4}
\parindent 0pt
%\floatstyle{boxed} 
\restylefloat{figure}
\usepackage[a4paper=true, ]{hyperref}

@@

\begin{figure}[b]
	\begin{center}
	\includegraphics[height=2cm,width=8cm]{ITC-logo.jpg}
	\end{center}
\end{figure}

@@

\vspace{4em}

\textbf{Software and utilities used in the current work}:\\
\begin{itemize}
	\item \href{http://www.esri.com/software/arcgis/index.html}{ArcGIS 10.0}: mapping, data integration and spatial analysis
	\item \href{http://www.erdas.com/products/ERDASIMAGINE/ERDASIMAGINE/Details.aspx}{Erdas Imagine}: imagery processing and classification
	\item \href{http://www.filewatcher.com/b/ftp/ftp.dfd.dlr.de/pub/WASI.0.0.html}{WASI Water Color Simulator}, for modelling optical water properties under changing environment
	\item \href{http://welcome.hp.com/country/us/en/prodserv/handheld.html}{iPAQ} and GPS for data capture during the fieldwork
	\item \href{http://www.bibtex.org/de/}{BibTeX}, for management of bibliography and reference database
	\item \href{http://www.latex-project.org/}{\LaTeX}, for document typesetting, formatting and performing pdf output
	\item \href{http://www.python.org/}{Python language}, for writing auxiliary scripts (e.g. raw data interpolation)
	\item \href{http://www.gnuplot.info/}{Gnuplot} for plotting graphs
	\item \href{http://www.openoffice.org/}{Open Office} : document typesetting, data preprocessing and preliminary statistical evalution
	\item \href{http://calc2latex.sourceforge.net/}{Calc2LaTeX} : converting spreadsheets into \LaTeX environment
	\item \href{http://www.spss.com/}{SPSS (PASW Statistics 18)} : statistical data edition and analysis 
	\item \href{http://gretl.sourceforge.net/}{Gretl: Gnu Regression, Econometrics and Time-series Library} for statistic analysis
	\item \href{http://staruml.sourceforge.net/en/}{StarUML} , \href{http://live.gnome.org/Dia}{Dia} , \href{http://argouml.tigris.org/}{ArgoUML}: diagramming flowcharts for work algorithms
	\item \href{http://inkscape.org/}{Inkscape}: drawing
\end{itemize}
\pagebreak

\section*{}
\markright{List of Figures}
\addcontentsline{toc}{section}{List of Figures}
\listoffigures
\pagebreak

@@

\pagenumbering{arabic}
\rfoot{MSc Thesis\\ \emph{"Seagrass mapping and monitoring along the coasts of Crete, Greece"}}
\section{Introduction}
%\markright{I. Introduction}

@@

The distribution of the seagrasses is strongly influensed by several environmental factors, which
include climate (mostly, tropical and temperate areas), bathymetry (shallow shelf zones), hydrological
particularities (chemical content of water, nutritient availability and turbidity of waves), and geological 
characteristics - sedimentation and cover types of the seafloor \cite{McKenzie06}\label{McKenzie06}.\\
There are four European seagrass species in Mediterranean area \cite{Borum04}\label{Borum04}: Zostera marina,
Zostera noltii, Cymodocea nodosa and \textit{P.oceanica}. In Greece the common species are\textit{P.oceanica}
(L.) Delile, Cymodocea nodosa (Ucria) Ascherson, Zostera noltii Hornemann and Halophila
stipulacea \cite{Amoutzopoulou-Schina05}\label{Amoutzopoulou-Schina05}. 
These species differ in morphological and
phenological features (Fig. \ref{fig:3}) as well as in structure and dynamics. Thus, Cymodocea nodosa is
considered the pioneer species of \textit{P.oceanica} beds, the latter species forming the last stage. When P.
oceanica beds regresses, C. nodosa often replaces them \cite{DenHartog77}\label{DenHartog77}; as a result, \textit{P.oceanica},
C. nodosa, and Z. noltii do not form mixed persistent stands \cite{Buia91}\label{Buia91}.

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.48]{UML_ClassDiagram.jpg}
	\caption{General methodological approach for the analysis of spectral signatures. \href{http://argouml.tigris.org/}{ArgoUML}}
	\label{fig:16}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.5]{UML_GIS.png}
	\caption{General methodological research approach. \href{http://inkscape.org/}{Inkscape}}
	\label{fig:16}
\end{figure}

\section[Overview of literature]{Seagrass monitoring: overview of literature and research resources} \label{sec:2}
%\markright{II. Seagrass monitoring: overview of literature and research resources}
%\renewcommand{\headrulewidth}{0.4pt}

@@

\begin{figure}
	\centering
	\subfloat [RAMSES-ACC-UV - Hyperspectral UVA/UVB Irradiance Sensor: 280-500 nm]{\label{fig:1-1a}\includegraphics[width=0.4\textwidth]{Fig-2-1.jpg}}
	\subfloat [RAMSES-ARC - Hyperspectral UV-VIS Radiance Sensor: 320-950 nm.]{\label{fig:1-1b}\includegraphics[width=0.4\textwidth]{Fig-2-2.jpg}}
	\caption{RAMSES Hyperspectral Sensors. Source: \href{http://www.trios.de}{Trios}}
	\label{fig:10}
\end{figure}

@@

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.30]{Fig-11.jpg}
	\caption{In situ optical reflectance spectra of seagrass. Shaded areas - percent of spectra lying within the range of reflectance. 
		White lines - mean spectra. Source: \cite{Hochberg03b}\label{Hochberg03b}}
	\label{fig:11}
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.25]{Fig-12.jpg}
	\caption{Difference between broaddand multispectral and hyperspectral resolution of
		spectral signatures. Source: \cite{Clark99}\label{Clark99}}
	\label{fig:12}
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.35]{Fig-13.jpg}
 	\caption{Band coverage of ASTER and Landsat channels on the e/m spectrum. Source:\cite{Kalinowski04}\label{Kalinowski04}}
	\label{fig:13}
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.30]{Fig-14.jpg}
	\caption{Multispectral vs. hyperspectral band coverage. Source: \cite{Sabins97}\label{Sabins97}}
	\label{fig:14}
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.15]{Fig-15.jpg}
	\caption{Reflectance spectra of seagrass; Rrs(0-) - subsurface RS reflectance; Rrs b - the bottom
		reflectance. Source:\cite{Yang10}\label{Yang10}}
	\label{fig:15}
\end{wrapfigure}

@@

\section{Materials and methods}
%\markright{III. Materials and methods}

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.10]{Fig-16.jpg}
	\caption{Study area: Crete island}
	\label{fig:16}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.25]{Fig-17.jpg}
	\caption{Study area: Ligaria beach, Crete island}
	\label{fig:17}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\subfloat [iPAQ, HP]{\label{fig:19a}\includegraphics[width=0.3\textwidth]{Fig-19a.jpg}}
	\subfloat [Waterproof Otterbox for iPAQ]{\label{fig:19a}\includegraphics[width=0.3\textwidth]{Fig-19b.jpg}}
	\subfloat [GPS]{\label{fig:19a}\includegraphics[width=0.15\textwidth]{Fig-19c.jpg}}
	\subfloat [Olympus waterproof camera]{\label{fig:19a}\includegraphics[width=0.25\textwidth]{Fig-19d.jpg}}
	\caption{Fieldwork equipment}
	\label{fig:18}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.70]{Fig-18.jpg}
	\caption{Boat used for the fieldwork measurements on Crete island. 2010}
	\label{fig:19}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\subfloat [view from below]{\label{fig:19a}\includegraphics[width=0.3\textwidth]{Fig-20a.jpg}}
	\subfloat [view \textit{en-face}]{\label{fig:19a}\includegraphics[width=0.3\textwidth]{Fig-20b.jpg}}
	\subfloat [view from above]{\label{fig:19a}\includegraphics[width=0.3\textwidth]{Fig-20c.jpg}}
	\caption{Scheme of placements of the Olimpus cameras during the measurements}
	\label{fig:20}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.40]{Fig-21.jpg}
	\caption{Locations of the video measurements and GPS tracklogs, Ligaria}
	\label{fig:21}
\end{figure}

@@

\subsection{Google Earth aerial imagery for seagrass mapping}
Apart from the satellite imagery, the aerial photographs from the Google Earth provide
a powerful tool for seagrass mapping, because they are important, reliable, detailed
and up-to-date source of imagery.

\begin{wrapfigure}{r}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.27]{Fig-22.jpg}
	\caption{Capturing aerial imagery from the Google Earth: grabbing process}
	\label{fig:22}
\end{wrapfigure}

Perhaps, the most clear advantage of the Google
Earth imagery is high resolution (15 m in land areas and lower in the oceans).
Obtained from the airborne platforms, Google Earth images have general spatial
resolution of several meters (though vatying in different areas), which allows very
detailed habitat and seafloor types discrimination, comparing with images
acquired from the spaceborne satellite platforms.
The spatial coverage of the Google imagery is lesser comparing to spaceborne images, but this can be solved as
well: while in general providing smaller area coverage than satellites images, the Google Earth
images can be stitched to the composite maps of the acceptable spatial extent, using Google grabbing
process which allows multiple overlapping of single images over the flight paths, and generates
mosaics.

@@

\pagebreak
\section{Results}
%\markright{IV. Results}

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.40]{UML_Dia.jpg}
	\caption{Workflow for the data aquisition, \href{http://live.gnome.org/Dia}{Dia} display}
	\label{fig:21}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.5]{GNU-17.jpg}
	\caption{Optical properties of the sea water with sediments, measured in aquarium tank. Agia Pelagia district, Crete. \href{http://www.gnuplot.info/}{Gnuplot}}
	\label{fig:23}
\end{figure}
\pagebreak

\subsection{Data pre-processing}
A collection of visible spectra of two seafloor cover types - \textit{P.oceanica} and carbonatic sand - consist of multiple measurement sets of P.oceanica made by means of RAMSES hyperspectral radiometer: 1-350 for 15th October, 1-445 for 14th October for P.oceanica (Tab. \ref{tab:10}); 1-106 for water without sedimnts, measured in aquarium tank, 1-27 for seawater with sediments measured in aquarium tank; 1-75 for carbonate sand (Tab. \ref{tab:17}). All these data have been analysed and statistically tested. We measured mean, extreme values (min-max), median and statistical quartiles, in order to visualise distribution of the values at varuious spectral wavelengths. Carbonate sand was measured at wavelengths 402-750 nm, while P.oceanica - at the interval 318-951 nm. We visualised the behavour of different spectra on relevant graphs (Fig. \ref{fig:43}, \ref{fig:44}, \ref{fig:45}). 
The available collected data were tested for the spectral variability and separability under varying conditions of dfifferent environmental
constituents (e.g. depth, water content, sun angle), in order to determine the
potential that it may have on the approaches for further images processing and classification. Behavour of the spectra of P.oceanica have been tested using datasets for various depths: 0.5, 1.5 and 2.5 meters (Tab. \ref{tab:8}, Tab. \ref{tab:9}). The raw
initial data of measurements of the spectral reflectance have been preprocessed and statistically analysed thereafter withgraphs visualising mean spectra and Q1-Q3 interval, instead of the series of single observations which is illustrated, for example, on Figures \ref{fig:27}, \ref{fig:33}, \ref{fig:28}). These data
have been measured using different step of the wavebands: some measurements
were made with 3 nm interval, while others – using 4 and 1 nm step.

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.30]{Fig-24.jpg}
	\caption{Script written on Python, for the interpolation raw data of the Trios-RAMSES measurements}
	\label{fig:24}
\end{figure}

Therefore, these data had to be interpolated and standardized to one
format, which is values of spectral reflectancese with one nm step. For the
interpolation we used script written on Python programming language that allowed to receive more 
detailed data by interpolating them from 3 and 4 nm step up to 2 nm (Fig.\ref{fig:24}).
The interpolated data contained spectral measurements of the seagrass P.oceanica (measured at Agia
Pelagia beach, Crete), sand (measured at Agia Pelagia beach, Crete), silt and default artificial
spectrum of constant albedo at WASI (Fig.\ref{fig:25}). 
We included measurements of the Thalassia seagrass
(measured at Southern Chinese Sea by C.Yang \cite{Yang10} \label{Yang10}) for the comparison of the spectral reflectances of
different seagrass speciea under various environmental conditions.
The most suitable wavebands for the seagrass monitroring usually lay between 400 and 700 nm,
which can be concluded by the visual examination, comparison and analysis of the different spectra of
the seagrasses. Therefore, we chosen the spectra 400-750 as the most appropriate range for
further research experiment.The results of the linear interpolation (Fig.\ref{fig:25}) demonstrate values of the
sand spectral reflectances with 1 nm interval covering the wavelength diapason of 400-750 nm.\\
The initial measured data were stored in raw-oriented format, so that re-formatting them into the
column-based layout was done using “transpose” command in Open Office or Excel (Fig. \ref{fig:42}). The next step included
calculation of the median, mean, quartiles and other statistical values at every data set (see Appendix, Fig. \ref{fig:54}, Fig. \ref{fig:55}).
After the preliminary analysis, the measured data were visualised using GNUplot programm, which
enables fine plotting of various datasets together: Fig \ref{fig:26}, Fig.\ref{fig:27}, Fig.\ref{fig:28}. The most acceptable method of interpolation was B\'ezier curve (Fig. \ref{fig:47}, Fig. \ref{fig:49}),
as it has trend-friendly graph better showing the general behavior of the curves at different wavebands
comparing to splines (see Appendix for more results).

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.25]{Fig-28.jpg}
	\caption{Fragment of \textit{Bottom.R} file: values of spectral measurements of the
		seagrasses (various species), sand, silt and artificial spectrum of constant albedo}
	\label{fig:25}
\end{figure}

Therefore, after several experiments with
various interpolation techniques (Fig.\ref{fig:48}), we have choosen Bézier curves interpolation (Fig.\ref{fig:49}) which contain convex
hulls made on its control points, and therefore is best suitable for our case: analysis of optical
properties of seawater.
%
\subsection{Analysis of spectral signatures}
The distinguishing spectral signatures for various seafloor types (e.g. seagrass species, coral reefs,
various types of sand, mud, other sediments, (Fig.\ref{fig:26})) exist in well-defined and narrow (10–20 nm)
wavelength ranges.

@@

\begin{figure}[h]
	\centering
	\subfloat [CZCS]{\label{fig:34a}\includegraphics[width=0.4\textwidth]{Fig-35a.jpg}}
	\subfloat [MERIS]{\label{fig:34b}\includegraphics[width=0.4\textwidth]{Fig-35b.jpg}}\\
	\subfloat [MODIS]{\label{fig:34c}\includegraphics[width=0.4\textwidth]{Fig-35c.jpg}}
	\subfloat [SeaWiFS]{\label{fig:34d}\includegraphics[width=0.4\textwidth]{Fig-35d.jpg}}
	\caption{Simulated remote sensing reflectance of \textit{P.oceanica} at various sensors}
	\label{fig:29}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\subfloat [depth of 0.5 m]{\label{fig:32a}\includegraphics[width=0.4\textwidth]{Fig-32a.jpg}}
	\subfloat [depth of 2.0 m]{\label{fig:32b}\includegraphics[width=0.4\textwidth]{Fig-32b.jpg}}\\
	\subfloat [depth of 3.5 m]{\label{fig:32c}\includegraphics[width=0.4\textwidth]{Fig-32c.jpg}}
	\subfloat [depths of 0.5, 2.0 and 3.5 m]{\label{fig:32d}\includegraphics[width=0.4\textwidth]{Fig-32d.jpg}}
	\caption{Remote sensing reflectance of \textit{P.oceanica} at various depths}
	\label{fig:32}
\end{figure}

>>>>

\documentclass[10pt, a4paper]{article}
\usepackage{color}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[numbers]{natbib}
\usepackage{calc}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage[svgnames,table]{xcolor}
\usepackage[subfigure]{tocloft}
%\usepackage[table]{xcolor}
\usepackage{paralist}
\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{fancyheadings}
\topmargin = 1pt
\headheight=10mm
\headsep=7pt
%\ensuremath
\usepackage{lastpage}
\usepackage{float}
\usepackage{subfig}
\usepackage{caption}
\usepackage{chngcntr}
\counterwithin{figure}{section}
\counterwithin{table}{section}
\usepackage{appendix}
\captionsetup{margin=10pt,font=small,labelfont=bf}
\usepackage{wrapfig}
\definecolor{darkblue}{rgb}{0.0,0.0,0.3}
\setcounter{tocdepth}{4}
\parindent 0pt
%\floatstyle{boxed} 
\restylefloat{figure}
\usepackage[a4paper=true, ]{hyperref}

@@

\begin{figure}[h]
	\begin{center}
	\includegraphics[height=2cm,width=8cm]{ITC-logo.jpg}
	\end{center}
\end{figure}

@@

\textbf{Software and utilities used in the current work}:\\
\begin{itemize}
	\item \href{http://www.esri.com/software/arcgis/index.html}{ArcGIS 10.0}: mapping, data integration and spatial analysis
	\item \href{http://www.erdas.com/products/ERDASIMAGINE/ERDASIMAGINE/Details.aspx}{Erdas Imagine}: imagery processing and classification
	\item \href{http://www.filewatcher.com/b/ftp/ftp.dfd.dlr.de/pub/WASI.0.0.html}{WASI Water Color Simulator}, for modelling optical water properties under changing environment
	\item \href{http://welcome.hp.com/country/us/en/prodserv/handheld.html}{iPAQ} and GPS for data capture during the fieldwork
	\item \href{http://www.bibtex.org/de/}{BibTeX}, for management of bibliography and reference database
	\item \href{http://www.latex-project.org/}{\LaTeX}, for document typesetting, formatting and performing pdf output
	\item \href{http://www.python.org/}{Python language}, for writing auxiliary scripts (e.g. raw data interpolation)
	\item \href{http://www.gnuplot.info/}{Gnuplot} for plotting graphs
	\item \href{http://www.openoffice.org/}{Open Office} : document typesetting, data preprocessing and preliminary statistical evalution
	\item \href{http://calc2latex.sourceforge.net/}{Calc2LaTeX} : converting spreadsheets into \LaTeX environment
	\item \href{http://www.spss.com/}{SPSS (PASW Statistics 18)} : statistical data edition and analysis 
	\item \href{http://gretl.sourceforge.net/}{Gretl: Gnu Regression, Econometrics and Time-series Library} for statistic analysis
	\item \href{http://staruml.sourceforge.net/en/}{StarUML} , \href{http://live.gnome.org/Dia}{Dia} , \href{http://argouml.tigris.org/}{ArgoUML}: diagramming flowcharts for work algorithms
	\item \href{http://inkscape.org/}{Inkscape}: drawing
	\item \href{http://www.notepad-plus-plus.org/}{Notepad++}: auxiliary text editing and raw data formatting
\end{itemize}
\pagebreak

\section*{}
\markright{List of Figures}
\addcontentsline{toc}{section}{List of Figures}
\listoffigures
\setcounter{lofdepth}{2}
\pagebreak

@@

\pagenumbering{arabic}
\rfoot{MSc Thesis\\ \emph{"Seagrass mapping and monitoring along the coasts of Crete, Greece"}}
\section{Introduction}
%%%%%%%%%%%%%%%%%%Section1%%%%%%%%%%%%%%%%%%
\counterwithin{figure}{section}
\counterwithin{table}{section}

\markright{I. Introduction}

@@

The distribution of the seagrasses is strongly influensed by several environmental factors, which
include climate (mostly, tropical and temperate are--as), bathymetry (shallow shelf zones), hydrological
particularities (chemical content of water, nutritient availability and turbidity of waves), and geological 
characteristics - sedimentation and cover types of the seafloor \cite{McKenzie06}\label{McKenzie06}.\\
There are four European seagrass species in Mediterranean area \cite{Borum04}\label{Borum04}: Zostera marina,
Zostera noltii, Cymodocea nodosa and \textit{P.oceanica}. In Greece the common species are\textit{P.oceanica}
(L.) Delile, Cymodocea nodosa (Ucria) Ascherson, Zostera noltii Hornemann and Halophila
stipulacea \cite{Amoutzopoulou-Schina05}\label{Amoutzopoulou-Schina05}. 
These species differ in morphological and
phenological features (Fig. \ref{fig:3}) as well as in structure and dynamics. Thus, Cymodocea nodosa is
considered the pioneer species of \textit{P.oceanica} beds, the latter species forming the last stage. When P.
oceanica beds regresses, C. nodosa often replaces them \cite{DenHartog77}\label{DenHartog77}; as a result, \textit{P.oceanica},
C. nodosa, and Z. noltii do not form mixed persistent stands \cite{Buia91}\label{Buia91}.

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.48]{UML_ClassDiagram.jpg}
	\caption{General methodological approach for the analysis of spectral signatures. \href{http://argouml.tigris.org/}{ArgoUML}}
	\label{fig:10}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.6]{UML_GIS.png}
	\caption{General methodological research approach. \href{http://inkscape.org/}{Inkscape}}
	\label{fig:11}
\end{figure}

\pagebreak
%%%%%%%%%%%%%%%%%Section2%%%%%%%%%%%%%%%%%%%%%%
\section[Overview of literature]{Seagrass monitoring: overview of literature and research resources} \label{sec:2}
\markright{II. Seagrass monitoring: overview of literature and research resources}
\renewcommand{\headrulewidth}{0.4pt}
\counterwithin{figure}{section}
\counterwithin{table}{section}

@@

\begin{figure}
	\centering
	\subfloat [RAMSES-ACC-UV - Hyperspectral UVA/UVB Irradiance Sensor: 280-500 nm]{\label{fig:1-1a}\includegraphics[width=0.4\textwidth]{Fig-2-1.jpg}}
	\subfloat [RAMSES-ARC - Hyperspectral UV-VIS Radiance Sensor: 320-950 nm.]{\label{fig:1-1b}\includegraphics[width=0.4\textwidth]{Fig-2-2.jpg}}
	\caption{RAMSES Hyperspectral Sensors. Source: \href{http://www.trios.de}{Trios}}
	\label{fig:12}
\end{figure}

@@

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.30]{Fig-11.jpg}
	\caption{In situ optical reflectance spectra of seagrass. Shaded areas - percent of spectra lying within the range of reflectance. 
		White lines - mean spectra. Source: \cite{Hochberg03b}\label{Hochberg03b}}
	\label{fig:13}
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.25]{Fig-12.jpg}
	\caption{Difference between broaddand multispectral and hyperspectral resolution of
		spectral signatures. Source: \cite{Clark99}\label{Clark99}}
	\label{fig:14}
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.35]{Fig-13.jpg}
 	\caption{Band coverage of ASTER and Landsat channels on the e/m spectrum. Source:\cite{Kalinowski04}\label{Kalinowski04}}
	\label{fig:15}
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.30]{Fig-14.jpg}
	\caption{Multispectral vs. hyperspectral band coverage. Source: \cite{Sabins97}\label{Sabins97}}
	\label{fig:16}
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.15]{Fig-15.jpg}
	\caption{Reflectance spectra of seagrass; Rrs(0-) - subsurface RS reflectance; Rrs b - the bottom
		reflectance. Source:\cite{Yang10}\label{Yang10}}
	\label{fig:17}
\end{wrapfigure}

@@

%%%%%%%%%%%%%%%%%  Section  3  %%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and methods}
\counterwithin{figure}{section}
\counterwithin{table}{section}

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.10]{Fig-16.jpg}
	\caption{Study area: Crete island}
	\label{fig:18}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.25]{Fig-17.jpg}
	\caption{Study area: Ligaria beach, Crete island}
	\label{fig:19}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\subfloat [iPAQ, HP]{\label{fig:20a}\includegraphics[width=0.3\textwidth]{Fig-19a.jpg}}
	\subfloat [Waterproof Otterbox for iPAQ]{\label{fig:20b}\includegraphics[width=0.3\textwidth]{Fig-19b.jpg}}
	\subfloat [GPS]{\label{fig:20c}\includegraphics[width=0.15\textwidth]{Fig-19c.jpg}}
	\subfloat [Olympus waterproof camera]{\label{fig:20d}\includegraphics[width=0.25\textwidth]{Fig-19d.jpg}}
	\caption{Fieldwork equipment}
	\label{fig:20}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.70]{Fig-18.jpg}
	\caption{Boat used for the fieldwork measurements on Crete island. 2010}
	\label{fig:21}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\subfloat [view from below]{\label{fig:22a}\includegraphics[width=0.3\textwidth]{Fig-20a.jpg}}
	\subfloat [view \textit{en-face}]{\label{fig:22a}\includegraphics[width=0.3\textwidth]{Fig-20b.jpg}}
	\subfloat [view from above]{\label{fig:22a}\includegraphics[width=0.3\textwidth]{Fig-20c.jpg}}
	\caption{Scheme of placements of the Olimpus cameras during the measurements}
	\label{fig:22}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.40]{Fig-21.jpg}
	\caption{Locations of the video measurements and GPS tracklogs, Ligaria}
	\label{fig:23}
\end{figure}

@@

\subsection{Google Earth aerial imagery for seagrass mapping}
Apart from the satellite imagery, the aerial photographs from the Google Earth provide
a powerful tool for seagrass mapping, because they are important, reliable, detailed
and up-to-date source of imagery.
Perhaps, the most clear advantage of the Google
Earth imagery is high resolution (15 m in land areas and lower in the oceans).
Obtained from the airborne platforms, Google Earth images have general spatial
resolution of several meters (though vatying in different areas), which allows very
detailed habitat and seafloor types discrimination, comparing with images
acquired from the spaceborne satellite platforms.
The spatial coverage of the Google imagery is lesser comparing to spaceborne images, but this can be solved as
well: while in general providing smaller area coverage than satellites images, the Google Earth
images can be stitched to the composite maps of the acceptable spatial extent, using Google grabbing
process which allows multiple overlapping of single images over the flight paths, and generates
mosaics.

@@

%%%%%%%%%%%%%%%%%  Section  4  %%%%%%%%%%%%%%%%%%%%%%%
\pagebreak
\section{Results}
\markright{IV. Results}
\counterwithin{figure}{section}
\counterwithin{table}{section}

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.40]{UML_Dia.jpg}
	\caption{Workflow for the data aquisition, \href{http://live.gnome.org/Dia}{Dia} display}
	\label{fig:24}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.5]{GNU-17.jpg}
	\caption{Optical properties of the sea water with sediments, measured in aquarium tank. Agia Pelagia district, Crete. \href{http://www.gnuplot.info/}{Gnuplot}}
	\label{fig:25}
\end{figure}
\pagebreak

\subsection{Data pre-processing}
A collection of visible spectra of two seafloor cover types - \textit{P.oceanica} and carbonatic sand - consist of multiple measurement sets of P.oceanica made by means of RAMSES hyperspectral radiometer: 1-350 for 15th October, 1-445 for 14th October for P.oceanica (Tab. \ref{tab:10}); 1-106 for water without sedimnts, measured in aquarium tank, 1-27 for seawater with sediments measured in aquarium tank; 1-75 for carbonate sand (Tab. \ref{tab:17}). All these data have been analysed and statistically tested. We measured mean, extreme values (min-max), median and statistical quartiles, in order to visualise distribution of the values at varuious spectral wavelengths. Carbonate sand was measured at wavelengths 402-750 nm, while P.oceanica - at the interval 318-951 nm. We visualised the behavour of different spectra on relevant graphs (Fig. \ref{fig:43}, \ref{fig:44}, \ref{fig:45}). 
The available collected data were tested for the spectral variability and separability under varying conditions of dfifferent environmental
constituents (e.g. depth, water content, sun angle), in order to determine the
potential that it may have on the approaches for further images processing and classification. Behavour of the spectra of P.oceanica have been tested using datasets for various depths: 0.5, 1.5 and 2.5 meters (Tab. \ref{tab:8}, Tab. \ref{tab:9}). The raw
initial data of measurements of the spectral reflectance have been preprocessed and statistically analysed thereafter withgraphs visualising mean spectra and Q1-Q3 interval, instead of the series of single observations which is illustrated, for example, on Figures \ref{fig:27}, \ref{fig:33}, \ref{fig:28}). These data
have been measured using different step of the wavebands: some measurements
were made with 3 nm interval, while others – using 4 and 1 nm step.
Therefore, these data had to be interpolated and standardized to one
format, which is values of spectral reflectancese with one nm step. For the
interpolation we used script written on Python programming language that allowed to receive more 
detailed data by interpolating them from 3 and 4 nm step up to 2 nm (Fig.\ref{fig:24}).
The interpolated data contained spectral measurements of the seagrass P.oceanica (measured at Agia
Pelagia beach, Crete), sand (measured at Agia Pelagia beach, Crete), silt and default artificial
spectrum of constant albedo at WASI (Fig.\ref{fig:25}). 
We included measurements of the Thalassia seagrass
(measured at Southern Chinese Sea by C.Yang \cite{Yang10} \label{Yang10}) for the comparison of the spectral reflectances of
different seagrass speciea under various environmental conditions.
The most suitable wavebands for the seagrass monitroring usually lay between 400 and 700 nm,
which can be concluded by the visual examination, comparison and analysis of the different spectra of
the seagrasses. Therefore, we chosen the spectra 400-750 as the most appropriate range for
further research experiment.The results of the linear interpolation (Fig.\ref{fig:25}) demonstrate values of the
sand spectral reflectances with 1 nm interval covering the wavelength diapason of 400-750 nm.\\
The initial measured data were stored in raw-oriented format, so that re-formatting them into the
column-based layout was done using “transpose” command in Open Office or Excel (Fig. \ref{fig:42}). The next step included
calculation of the median, mean, quartiles and other statistical values at every data set (see Appendix, Fig. \ref{fig:54}, Fig. \ref{fig:55}).
After the preliminary analysis, the measured data were visualised using GNUplot programm, which
enables fine plotting of various datasets together: Fig \ref{fig:26}, Fig.\ref{fig:27}, Fig.\ref{fig:28}. The most acceptable method of interpolation was B\'ezier curve (Fig. \ref{fig:47}, Fig. \ref{fig:49}),
as it has trend-friendly graph better showing the general behavior of the curves at different wavebands
comparing to splines (see Appendix for more results). 
Therefore, after several experiments with
various interpolation techniques (Fig.\ref{fig:48}), we have choosen Bézier curves interpolation (Fig.\ref{fig:49}) which contain convex
hulls made on its control points, and therefore is best suitable for our case: analysis of optical
properties of seawater.
%
\subsection{Analysis of spectral signatures}
The distinguishing spectral signatures for various seafloor types (e.g. seagrass species, coral reefs,
various types of sand, mud, other sediments, (Fig.\ref{fig:26})) exist in well-defined and narrow (10–20 nm)
wavelength ranges.

@@

\begin{figure}[h]
	\centering
	\subfloat [CZCS]{\label{fig:29a}\includegraphics[width=0.4\textwidth]{Fig-35a.jpg}}
	\subfloat [MERIS]{\label{fig:29b}\includegraphics[width=0.4\textwidth]{Fig-35b.jpg}}\\
	\subfloat [MODIS]{\label{fig:29c}\includegraphics[width=0.4\textwidth]{Fig-35c.jpg}}
	\subfloat [SeaWiFS]{\label{fig:29d}\includegraphics[width=0.4\textwidth]{Fig-35d.jpg}}
	\caption{Simulated remote sensing reflectance of \textit{P.oceanica} at various sensors}
	\label{fig:29}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\subfloat [depth of 0.5 m]{\label{fig:32a}\includegraphics[width=0.4\textwidth]{Fig-32a.jpg}}
	\subfloat [depth of 2.0 m]{\label{fig:32b}\includegraphics[width=0.4\textwidth]{Fig-32b.jpg}}\\
	\subfloat [depth of 3.5 m]{\label{fig:32c}\includegraphics[width=0.4\textwidth]{Fig-32c.jpg}}
	\subfloat [depths of 0.5, 2.0 and 3.5 m]{\label{fig:32d}\includegraphics[width=0.4\textwidth]{Fig-32d.jpg}}
	\caption{Remote sensing reflectance of \textit{P.oceanica} at various depths}
	\label{fig:32}
\end{figure}
